The Neuro Bureau ADHD-200 Preprocessed Repository

### 2016-05-28 responses_reviewers.tex

\ead{pierre.bellec@criugm.qc.ca}
\ead[url]{bellec.simexp-lab.org}

\author[label0,label3]{Carlton Chu}
\address[label3]{Google DeepMind, London, UK}
\ead{carltonchu1@gmail.com}

\author[label0,label1,label4]{Fran\c{c}ois Chouinard-Decorte}
\address[label4]{Integrated Program in Neuroscience, McGill University, Montreal, CA}
\ead{francois.chouinard@gmail.com}

\author[label0,label1,label5]{Yassine Benhajali}
\address[label5]{D\'epartement d'Anthropologie, Universit\'e de Montr\'eal, Montr\'eal, CA}
\ead{yanamarji@gmail.com}

\author[label0,label6]{Daniel S. Margulies}
\address[label6]{Max Planck Research Group for Neuroanatomy \& Connectivity, Max Planck Institute for Human Cognitive and Brain Sciences, Leipzig, Germany}
\ead{margulies@cbs.mpg.de}

\author[label0,lab7,lab8]{R. Cameron Craddock\corref{cor1}}
\address[lab7]{Computational Neuroimaging Laboratory, Center for Biomedical Imaging and Neuromodulation, Nathan S. Kline Institute for Psychiatric Research, Orangeburg, NY, USA}
\address[lab8]{Center for the Developing Brain, Child Mind Institute, New York, NY, USA}
\ead{ccraddock@nki.rfmh.org}
\ead[url]{computational-neuroimaging-lab.org}

### 2016-05-28 Bellec_ADHD200_2016.tex
Updates from ShareLaTeX

With the exception of a few datasets that were missing data, and one dataset that was corrupted, all of the ADHD-200 Sample was processed and released by Athena, regardless of data quality. This was done to accomodate differing opinions as to what qualifies as usable data, and to provide poor quality data that may be used by others to develop methods that are robust to noise. Files containing the six motion parameters for each rs-fMRI data and anatomical and mean EPI images in template space have been included in the release to enable users to determine high-motion data or poor registrations. 
\par
An online report\footnote{\url{http://simexp.github.io/adhd200_qc_athena/index.html}} of registration quality was generated for all subjects. The report includes concurrent visualization of the individual structural images, after non-linear registration, together with the target brain template (MNI pediatric), as well as a concurrent visualization of the individual structural and BOLD images together, again after non-linear registration. All these images were visually reviewed, and registration that presented severe misregistration or image quality problems were marked as ``fail'' for QC ($7\%$ failure rate). The outcome of the visual quality assessment for registered images has been shared on NITRC in a consolidated tabular-separated (.tsv) spreadsheet\footnote{\url{https://www.nitrc.org/frs/download.php/9024/adhd200_preprocessed_phenotypics.tsv}} that also includes all phenotypic variables. The .tsv QC spreadsheet also includes a summary measure of motion amplitude for each dataset, that can be used to exclude subjects above a certain threshold. 
\par
Additional quality metrics derived from the preprocessed s-MRI, GM masks, mean rs-fMRI images, fALFF maps, and FC maps were also included to help with the QC process. For each data type, registered in stereotaxic space,  a mean and standard deviation image was calculated from all of the scans from all of the subjects. These images were used to perform a voxelwise z-score transformation on each data type for each subject $z_i = \frac{(v_i-m_i)}{\sigma_i}$. The absolute values of these z-scores were thresholded at 3, and summed across voxels for a quality score for that image. The higher the resulting sum, the larger the number of voxels a image has that are $\geq 3$ and the more likely the images are outliers. Using this metric it is possible to rank images, and allows for a more directed search for poor quality scans. These metrics are distributed in participant-specific text files along with the BASH scripts that implement the procedure.

@@

In line with the original purpose of the ADHD-200 Global competition, most works derived from the ADHD-200 Preprocessed initiative proposed new machine learning techniques and their main application on real data was to predict clinical diagnosis from the ADHD-200 Preprocessed sample. The most frequently used derivatives have been Athena preprocessed fMRI volumes \citep[e.g.][]{Liang2012,Carmona2015} and Athena regional time series \citep[e.g.][]{Fujita2013,Dey2014}. Some works also based their main analysis on the Burner, either in isolation \citep[e.g.][]{Rangarajan2014,Rangarajan2015,Mahanand2013} or in combination with the Athena functional derivatives \citep[e.g.][]{Ji2011,Eloyan2012,Olivetti2014}. A few papers utilized the Athena ICA, ReHo and fALFF maps \citep[e.g.][]{Sato2012a} and the NIAK high resolution (either ROI1000 or ROI3000) regional time series in their main analysis \citep[e.g.][]{Ahn2015,Yang2015}. Some publications also made a more marginal use of the ADHD-200 Preprocessed repository, by re-using the CC200 or CC400 functional brain parcellations \citep[e.g.][]{Dey2012,Dai2012}, and not the sample itself. 

@@

To the best of our knowledge, no published works have tried to directly compare the results generated by the Athena and NIAK pipeline. The two pipelines conceptually implement very similar steps, with key operations including the non-linear volumetric registration of individual structural MRIs in two different variants of the MNI space, and the registration of invidual BOLD and T1 images. The most striking difference lied in the reduction of structured noise: the Athena implemented a regression of a battery of nuisance variables (motion parameters, WM and GM signals), while NIAK implemented an automated labeling of structured noise in a spatial ICA \citep{perlbarg2007corsica}. The Athena strategy is fairly standard in the rs-fMRI community, while ICA-based noise attenuation techniques are less common. The other clear difference lies in the derivatives offered by the two pipelines: Athena featured a more extensive battery of maps and regional time series, while the NIAK included high-resolution (several thousands) brain parcels and associated time series. Beyond personal ideology, or the availability of particular features, there is no clear reason to prefer one pipeline over the other.  It has recently become clear that even subtle variations of processing environments can impact the end-results of processing pipelines \citep{Glatard2015-we}, even when the same processing is replicated. We thus believe it is important to check how variable conclusions drawn from MRI data are, depending of the specific software implementation used for processing. The NIAK and Athena pipeline will enable researchers to compare the robustness of their tools or analysis results to different processing choices and derivatives. 

@@

The two biggest challenges for using the ADHD-200 Preprocessed data are head motion \cite{power2012, vandijk2012, sattertwhaite2012, fair2012adhd200, yan2013motion, yan2013small} and inter-site variation in the acquisition equipment, parameters, and experimental procedures \cite{Olivetti2012,yan2013std}. A variety of different approaches have been proposed for addressing head motion in hyperkinetic populations \cite{power2012, sattertwhaite2013}, and in the ADHD-200 Sample in particular \cite{fair2012adhd200}, that should be considered when analyzing the data. At the very least, some statistic that characterizes individual motion (such as root mean square deviation \cite{JenkinsonTR99}) should be included as a nuisance regressor in the group-level model \cite{sattertwhaite2012, yan2013motion}. Differences in the manner in which data was collected at each site can introduce addition and multiplicative effects (batch effects) to the data, which may obscure the underlying biological signal \cite{Olivetti2012,yan2013std}. Including a regressor for acquisition protocol (see Tables~\ref{smri_params_table} and \ref{rsfmri_params_table} for a summary of the different protocols), the average pairwise correlation between all regions in the brain (GCOR) \cite{saad2013gcor}, or the whole-brain average of the feature under inquiry \cite{yan2013std}, have all been shown to be effective for dealing with inter-site variation.

@@

The ADHD-200 Preprocessed initiative was successful in terms of its primary objectives: the derivatives shared in the repository were effectively used by many researchers during and after the ADHD-200 Global Competition, with over 10,500 downloads by more than 600 users, as well as 52 resulting publications \cite{Rangarajan2014, Liang2012, Tabas2014, Rangarajan2015, Mahanand2013, Lifshitz2012, Fujita2013, Ji2011, Li2015, Li2013, Liu2012, DosSantosSiqueira2014, Olivetti2014, Han2015, Wang2013a, Dey2014, Bellec2012, Bohland2012, Chang2012, Cheng2012, Colby2012, Dai2012, Dey2012, Eloyan2012, Olivetti2012, Sato2012a, Subramanian2013, Carmona2015, Hou2015, Deshpande2015, She2014, Lavoie-Courchesne2012b, Chen2015, Nunez-Garcia2015, Solmaz2012, Anderson2014, KadkhodaeianBakhtiari2012, Sato2013, Kyeong2015, Sato2012, Takahashi2012, He2013, Kong2013, Yao2013, Yang2015, Ahn2015, Fujita2014, Reiss2014,nachamai2016sub,Brown20161238,Yu2016,Fujita2015}, 3 PhD theses \cite{Colby2012a, Dey2014phd, Zhang2012}, 3 master's dissertations \cite{VanGalenLast2011, Vidal2014, Wang2013}, and 1 patent \cite{Dey2013} derived from the release in just over three years. Further publications are either in press or under review, and an updated list of publications will be maintained on the Mendeley platform\footnote{\url{https://www.mendeley.com/groups/4198361/adhd-200-preprocessed/}}. Although there was clearly a peak in usage around the ADHD-200 Global competition, there has been a sustained ammount of downloads and publications since then, see Figure \ref{fig:usage}, which we take as a demonstration of a long-term interest from the community in this resource. 

@@

The ADHD-200 preprocessed has helped to expand the boundaries of the traditional neuroimaging community, with several publications in core engineering and statistics journals that do not routinely feature neuroimaging applications, such as Statistica Sinica \citep{Yang2015} or the Journal of the American Statistical Association \citep{Chen2015}, amongst others \citep[e.g.][]{Liang2012, Rangarajan2015, Rangarajan2014, Mahanand2013, Ji2011, Li2015, Li2013, Liu2012, Subramanian2013, Hou2015, Deshpande2015, She2014, He2013, Kong2013, Ahn2015}. In particular, the winning team of the ADHD-200 Global Competition was based at the Johns Hopkins Biostatistics Department and used ADHD-200 Preprocessed to develop their diagnostic algorithm \cite{Eloyan2012}. We also found interesting that a handful of methodological works used the ADHD-200 Preprocessed sample as one application in a series of benchmarks that could be as varied as positron emission tomography in Alzheimer's disease \citep{Yang2015}, the Yeast gene regulatory network \citep{She2014} or gene expression in brain tissues of patients with HIV-1 associated neurocognitive disorders \citep{Liu2012}. This suggests that fully processed, easy-to-access imaging samples could help validate general-purpose methods on a wider scope of applications.

@@

The impact of the ADHD-200 Preprocessed repository demonstrated the need for reducing computational barriers to participation in discovery neuroscience, including but not limited to machine learning competitions based on neuroimaging data. The ADHD-200 Preprocessed release has contributed and will continue to contribute to respond to that need. We still believe that a much larger-scale effort will be necessary to unlock the full potential of openly shared neuroimaging data in the service of accelerating neuroimaging research. In line with the grassroots, open science ethos of the Neurobureau, new contributors interested in sharing derivatives of ADHD-200 or other open imaging data repository can contact us via a forum\footnote{\url{http://www.nitrc.org/forum/?group_id=383}}. We are also actively seeking for contributions, notably during brainhack events\citep{Cameron_Craddock2016-du}. An important current project is the ABIDE preprocessed initiative\footnote{\url{http://preprocessed-connectomes-project.org/abide/}} \citep{dimartino2014}. This new resource, still under development, will include over 15 different processing strategies for rs-fMRI processing, implemented across 4 different software packages, and 2 different pipelines for structural MRI processing. The release will feature an harmonized organization of processed derivatives across packages and extensive quality control. We are planning to continue to expand this line on work on other data sources in the near future. Our hope is that ADHD-200 Preprocessed and future related efforts will critically help fMRI researchers to identify optimal analytical paths for a given task.

>>>>

Images were visually inspected and attempts were made to fix gross misregistration errors by hand adjusting the offending images to center them on the anterior commisure and rotate them into rough correspondence with the MNI template. With the exception of a few datasets that were missing data, and one dataset that was corrupted, all of the ADHD-200 Sample was processed and released by Athena, regardless of data quality. This was done to accommodate differing opinions as to what qualifies as usable data, and to provide poor quality data that may be used by others to develop methods that are robust to noise. Files containing the six motion parameters for each rs-fMRI data and anatomical and mean EPI images in template space have been included in the release to enable users to determine high-motion data or poor registrations.
\par
Additional quality metrics derived from the preprocessed s-MRI, GM masks, mean rs-fMRI images, fALFF maps, and FC maps were also included to help with the QC process. For each data type a mean and standard deviation image was calculated from all of the scans (in stereotaxic space) from all of the subjects. These images were used to perform a voxelwise z-score transformation on each data type for each subject $z_i = \frac{(v_i-m_i)}{\sigma_i}$. For each image (or map), the number of these z-scores whose absolute value exceeded 3 where summed to generate a quality score. The higher the resulting sum, the larger the number of voxels a image has with $\left|z\right| \geq 3$ and the more likely the images are outliers. Using this metric it is possible to rank images for a more directed search for poor quality scans. These metrics are distributed in participant-specific text files along with the BASH scripts that implement the procedure.
\par
A rigorous visual inspection of registration has also been performed using the same procedures and raters (PB and YB) as the NIAK pipeline. The procedure entailed generating an online report\footnote{\url{http://simexp.github.io/adhd200_qc_athena/index.html}} of registration quality for all subjects.  The report includes a visualization of the individual structural images in standard space overlaid onto the corresponding brain template (MNI pediatric), as well as a visualization of the individual mean EPI image overlaid onto the corresponding structural image, both in standard space. These images were visually reviewed, and images with severe misregistraion or quality problems were marked as ``fail'' ($7\%$ failure rate). The outcome of the visual quality assessment along with summary measures of motion for the resting state fMRI scans and all other phenotypic variables have been consolidated into a tabular-separated (.tsv) spreadsheet that can be downloaded from NITRC\footnote{\url{https://www.nitrc.org/frs/download.php/9024/adhd200_preprocessed_phenotypics.tsv}}.

@@

In line with the original purpose of the ADHD-200 Global competition, most of the publications that have used the ADHD-200 Preprocessed initiative data have proposed new machine learning techniques for predicting ADHD diagnosis or subtype. The most frequently used derivatives have been Athena preprocessed fMRI volumes \citep[e.g.][]{Liang2012,Carmona2015} and Athena regional time series \citep[e.g.][]{Fujita2013,Dey2014}. Works have used Burner pipeline data both in isolation \citep[e.g.][]{Rangarajan2014,Rangarajan2015,Mahanand2013} or in combination with the Athena functional derivatives \citep[e.g.][]{Ji2011,Eloyan2012,Olivetti2014}. A fewer number of publications have also used Athena ICN, ReHo and fALFF maps \citep[e.g.][]{Sato2012a} and the NIAK high resolution (either ROI1000 or ROI3000) regional time series in their main analysis \citep[e.g.][]{Ahn2015,Yang2015}. Researchers have also found value in this initiative beyond the processed data, such as the CC200 or CC400 functional brain parcellations \citep[e.g.][]{Dey2012,Dai2012} and the Athena processing scripts \citep[.e.g][]{Fujita2015}. The three PhD dissertations \cite{Colby2012a, Dey2014phd, Zhang2012} and three master's theses \citep[][]{VanGalenLast2011,Wang2013,Vidal2014} that have used this resource have all focused on disease state prediction and data dimensionality reduction techniques.

@@

To the best of our knowledge, there are no published comparisons between the results generated by the Athena and NIAK pipeline. The two pipelines conceptually implement very similar steps, with key operations including the non-linear volumetric registration of individual structural MRIs in two different variants of the MNI space, and the registration of individual EPI and T1 images. The most striking differences between the pipelines are in the software that they used and the opinions that drove the parameter choices. The Athena implemented nuisance variable regression \citep{lund2006_nvr,fox2005} (motion parameters, WM and GM signals), while NIAK implemented an automated labeling of structured noise in a spatial ICA \citep{perlbarg2007corsica}. The Athena strategy is fairly standard in the rs-fMRI community, while ICA-based noise attenuation techniques are less common. Athena provided time series for a variety of low-dimension anatomical and functional brain parcellations, whereas NIAK favored much higher-resolution (several thousands) brain parcels. There are also differences in the type of data that each pipeline offers, Athena provides a wide variety of statistical derivatives calculated from both the functional and structural data, whereas Burner provided structural derivatives only, and NIAK primarily focused on function. Beyond personal ideology, the availability of particular features, or allegiances to certain software tools there is no clear reason to prefer one pipeline over the other.  It has recently become clear that even subtle variations of processing environments can impact the end-results of processing pipelines \citep{Glatard2015-we}, even when the same processing is replicated. The variety of data available through this initiative enables researchers to compare the robustness of their tools or analysis results to different processing choices and derivatives. 

@@

The two biggest challenges for using the ADHD-200 Preprocessed data are head motion \cite{power2012, vandijk2012, sattertwhaite2012, fair2012adhd200, yan2013motion, yan2013small} and inter-site variation in the acquisition equipment, parameters, and experimental procedures \cite{Olivetti2012,yan2013std}. A variety of different approaches have been proposed for addressing head motion in hyperkinetic populations \cite{power2012, sattertwhaite2013}, and in the ADHD-200 Sample in particular \cite{fair2012adhd200}, that should be considered when analyzing the data. At the very least, some statistic that characterizes individual motion (such as root mean square deviation \cite{JenkinsonTR99}) should be included as a nuisance regressor in the group-level model \cite{sattertwhaite2012, yan2013motion}. Differences in the manner in which data was collected at each site can introduce additive and multiplicative effects (batch effects) to the data, which may obscure the underlying biological signal \cite{Olivetti2012,yan2013std}. Including a regressor for acquisition protocol (see Tables~\ref{smri_params_table} and \ref{rsfmri_params_table} for a summary of the different protocols), the average pairwise correlation between all regions in the brain (GCOR) \cite{saad2013gcor}, or the whole-brain average of the feature under inquiry \cite{yan2013std}, have all been shown to be effective for dealing with inter-site variation.

@@

The ADHD-200 Preprocessed initiative was successful in terms of its primary objectives: the derivatives shared in the repository were effectively used by many researchers during and after the ADHD-200 Global Competition, with over 10,500 downloads by more than 600 users, as well as 52 resulting publications \cite{Rangarajan2014, Liang2012, Tabas2014, Rangarajan2015, Mahanand2013, Lifshitz2012, Fujita2013, Ji2011, Li2015, Li2013, Liu2012, DosSantosSiqueira2014, Olivetti2014, Han2015, Wang2013a, Dey2014, Bellec2012, Bohland2012, Chang2012, Cheng2012, Colby2012, Dai2012, Dey2012, Eloyan2012, Olivetti2012, Sato2012a, Subramanian2013, Carmona2015, Hou2015, Deshpande2015, She2014, Lavoie-Courchesne2012b, Chen2015, Nunez-Garcia2015, Solmaz2012, Anderson2014, KadkhodaeianBakhtiari2012, Sato2013, Kyeong2015, Sato2012, Takahashi2012, He2013, Kong2013, Yao2013, Yang2015, Ahn2015, Fujita2014, Reiss2014,nachamai2016sub,Brown20161238,Yu2016,Fujita2015}, 3 PhD theses \cite{Colby2012a, Dey2014phd, Zhang2012}, 3 master's dissertations \cite{VanGalenLast2011, Vidal2014, Wang2013}, and 1 patent \cite{Dey2013} derived from the release in just over three years. Further publications are either in press or under review, and an updated list of publications will be maintained in a public Mendeley group\footnote{\url{https://www.mendeley.com/groups/4198361/adhd-200-preprocessed/}}. Although there was clearly a peak in usage around the ADHD-200 Global competition, there has been a sustained amount of downloads and publications since then, see Figure \ref{fig:usage}, which we take as a demonstration of a long-term interest from the community in this resource. 

@@

The ADHD-200 preprocessed has helped to expand the boundaries of the traditional neuroimaging community, with several publications in core engineering and statistics journals that do not routinely feature neuroimaging applications, such as Statistica Sinica \citep{Yang2015} or the Journal of the American Statistical Association \citep{Chen2015}, amongst others \citep[e.g.][]{Liang2012, Rangarajan2015, Rangarajan2014, Mahanand2013, Ji2011, Li2015, Li2013, Liu2012, Subramanian2013, Hou2015, Deshpande2015, She2014, He2013, Kong2013, Ahn2015}. In particular, the winning team of the ADHD-200 Global Competition was based at the Johns Hopkins Biostatistics Department and used ADHD-200 Preprocessed to develop their diagnostic algorithm \cite{Eloyan2012}. We also found interesting that a handful of methodological publications used the ADHD-200 Preprocessed sample as one application in a series of benchmarks that could be as varied as positron emission tomography in Alzheimer's disease \citep{Yang2015}, the Yeast gene regulatory network \citep{She2014} or gene expression in brain tissues of patients with HIV-1 associated neurocognitive disorders \citep{Liu2012}. This suggests that fully processed, easy-to-access imaging samples could help validate general-purpose methods on a wider scope of applications.

@@

The impact of the ADHD-200 Preprocessed repository demonstrated the need for reducing computational barriers to participation in discovery neuroscience, including but not limited to machine learning competitions based on neuroimaging data. While the ADHD-200 Preprocessed initiative will have a long-term impact on that need, we believe that a much larger-scale effort will be necessary to unlock the full potential of openly shared neuroimaging data in the service of accelerating neuroscience research. In line with the grassroots, open science ethos of the Neurobureau, new contributors interested in sharing derivatives of ADHD-200 or other open imaging data repository can contact us via our web-based forum\footnote{\url{http://www.nitrc.org/forum/?group_id=383}}. We are  actively seeking new contributions to the Preprocessed Connectomes Project, notably during Brainhack events\citep{Cameron_Craddock2016-du}. An important current project is the ABIDE preprocessed initiative\footnote{\url{http://preprocessed-connectomes-project.org/abide/}} \citep{dimartino2014}. This new resource, still under development, will include 16 different processing strategies for rs-fMRI processing, implemented across 4 different software packages, and 2 different pipelines for structural MRI processing. The release will feature an harmonized organization of processed derivatives across packages and extensive quality control. We are planning to continue to expand this line on work on other data sources in the near future. Our hope is that ADHD-200 Preprocessed and future related efforts will critically help fMRI researchers to identify optimal analytical paths for a given task.
